Pattern matching for $(231,213)$-avoiding permutations

### 2015-01-30 article.tex
LNCS format

\documentclass[12pt, a4paper]{article}

\usepackage{amsfonts}
\usepackage{amsmath}

\usepackage[all]{xy}

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

@@

\title{Pattern matching for (231,213) avoiding permutations}
\author{Both Emerite NEOU, Romeo RIZZI}
\date{}

\begin{document}
	\maketitle 
	\tableofcontents

	\abstract{
	  	Given two permutations, 
	  	a text $\ptext$ 
	  	and a motif $\pmotif$,
	  	the \textbf{permutation pattern problem}
	  	asks if there exists a normalized subsequence of $\ptext$
	  	which equals $\pmotif$.
	  	We study this problem for the permutation 
	  	avoiding (231,231).
	}

	\section{Introduction}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[a4paper]{llncs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[all]{xy}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{complexity}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\DeclareMathOperator{\RUN}{run}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Pattern matching for $(231,213)$-avoiding permutations}
\author{Both Emerite NEOU, Romeo RIZZI}
\date{}

\author{%
	Both Emerite Neou\inst{1,3}
  Romeo Rizzi\inst{2} \and
  St\'ephane Vialette\inst{3}
}% end author
\institute{%
  Department of Computer Science,
  Università degli Studi di Trento, Italy \\	
	\and
  Department of Computer Science,
  Università degli Studi di Verona, Italy \\
  \email{romeo.rizzi@univr.it}
  \and
	Universit\'e' Paris-Est, LIGM (UMR 8049), CNRS, UPEM, ESIEE Paris, ENPC, 
	F-77454, Marne-la-Vallée, France\\
  \email{vialette@univ-mY.fr}
}% end institute

\date{\today}

\maketitle

\tableofcontents

\begin{abstract}
bla bla\ldots
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\label{section:Introduction}

	A permutation $\pi$ is said to contain another permutation $\sigma$,
	in symbols $\sigma \preceq \pi$,
	if there exists a subsequence of entries of $\pi$ that has the same relative
	order as $\sigma$, and in this case $\sigma$ is said to be a 
	\emph{pattern} of $\pi$.
	Otherwise, $\pi$ is said to avoid the permutation $\sigma$.
	For example a permutation contains the pattern $123$ (resp. $321$) if it has
	an increasing (resp. decreasing) subsequence of length $3$.
	Here, note that members need not actually to be consecutive, 
	merely increasing (resp. descreasing).
	During the last decade, the study of the pattern containment on permutations has
	become a very active area of research.

	We consider here the so-called \emph{pattern containment} problem
	(also sometimes referred to as the pattern involvement problem):
	Given two permutations $\sigma$ and $\pi$, this problem is to decide whether
	$\sigma \preceq \pi$ (the problem is ascribed to Wilf in \cite{Bose:Buss:Lubiw:1998}).
	The permutation containment problem is \NPhard~\cite{Bose:Buss:Lubiw:1998}.
	It is, however, polynomial time solvable by brute-force enumeration
	if $\sigma$ has bounded size.
	Improvements to this algorithm were presented in
	\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001} and
	\cite{Ahal:Rabinovich:2008},
	the latter describing a $O(|\pi|^{0.47k+o(|\sigma| )})$ time algorithm.
	Bruner and Lackner \cite{Bruner:Lackner:SWAT:2012}
	gave a fixed-parameter algorithm solving the pattern containment problem with
	an exponential worst-case runtime of $O(1.79^{\RUN(\pi)})$,
	where $\RUN(\pi)$ denotes the number of alternating runs of $\pi$.
	(This is an improvement upon the $O(2^{|\pi|})$ runtime required by
	brute-force search without imposing restrictions on $\sigma$ and $\pi$.)

	A few particular cases of the pattern containment problem have been attacked successfully.
	The case of increasing patterns is solvable in
	$O(|\pi| \log \log |\sigma|)$ time in the RAM model \cite{Crochemore:Porat:2010},
	improving the previous 30-year bound of $O(|\pi| \log |\sigma|)$.
	(The algorithm also improves on the previous
	$O(|\pi| \log \log |\pi|)$ bound.)
	Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear time
	by stack sorting algorithms.
	Any pattern of length $4$ can be detected in $O(|\pi| \log |\pi|)$ time
	\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
	Algorithmic issues for $321$-avoiding patterns containment has been investigated in
	\cite{Guillemot:Vialette:ISAAC:2009}.

	Of particular interest in our context,
	the pattern containment problem is also solvable in
	polynomial time for separable patterns.
	Separable permutations are those permutations that contain neither
	$2413$ nor $3142$, and they are enumerated by the Schröder numbers
	(sequence A006318 in OEIS).
	Separable permutations has gained a lot of interest in the pattern containment
	context during the last years.
	Indeed, the pattern containment problem is solvable $O(kn^4)$ time and
	$O(kn^3)$ space for separable patterns \cite{Ibarra:1997}
	(see also \cite{Bose:Buss:Lubiw:1998}),
	where $k$ is the length of the pattern and $n$ is the length of the
	target permutation.
	Notice that there are numerous characterizations of separable permutations.
	To mention just a few examples,
	they are the permutations whose permutation graphs are cographs
	(\emph{i.e.} $P_4$-free graphs);
	equivalently, a separable permutation is a permutation that can be obtained
	from the trivial permutation $1$ by \emph{direct sums} and \emph{skew sums}.
	To the best of our knowledge,
	separable permutations first arose in the work of
	Avis and Newborn~\cite{Avis:Monroe:1981},
	who showed that they are precisely the permutations which can be sorted by an
	arbitrary number of pop-stacks in series,
	where a pop-stack is a restricted form of stack in which any pop operation
	pops all items at once.
	(Notice that the separable permutations include as a special case the
	stack-sortable permutations, which avoid the pattern $231$.)
	Separable permutations
	have been widely studied in the last decade, both from a combinatorial
	\cite{West:1995} and an algorithmic
	\cite{Ibarra:1997,Bose:Buss:Lubiw:1998,Rossin:Bouvel:06,Bouvel:Rossin:Vialette:CPM:2007} point
	of view.

	There exist many generalisation of patterns that are worth considering
	in the context of algorithmic issues in pattern involvement
	(see \cite{Kitaev:book:2011} for an up-to-date survey).
	Vincular patterns, also called generalized patterns,
	resemble (classical) patterns, with the constraint that some of the letters in
	an occurrence must be consecutive.
	Of particular importance in our context,
	Bruner and Lackner \cite{Bruner:Lackner:SWAT:2012}
	proved that deciding whether a vincular pattern
	$\sigma$ of length $k$ occurs in a permutation
	$\pi \in \mathfrak{S}_n$ is $W[1]$-complete for
	parameter $k$.
	bivincular patterns generalize classical patterns even further than vincular
	patterns. Indeed, in bivincular patterns, not only positions but also values
	of elements involved in a matching may be forced to be adjacent
